Optimal Finite Element Integration

### 2015-08-06 TOMS_Paper/assembly.tex
Several minor fixes in perf eval section

\subsection{Performance of Forms}
\label{sec:perf-results-forms}
The first, general observation is that optimality, as we defined it, does not necessarily imply best achievable performance. There are two reasons for this. The first reason is that low-level optimizations can impact the actual runtime. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} tend to slightly outperform \texttt{cfO2} even though pre-evaluation is triggered. With $p=4$, \texttt{tens} is quite faster then \texttt{cfO2}; in particular, in this case \texttt{cfO2} considers pre-evaluation unprofitable. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate any operations involving zero-valued entries and 2) can drastically decrease the working set by not having to store entire temporaries. In COFFEE we never unroll loops to maximize the chances of autovectorization, which results in considerable speed ups throughout the majority of test cases. The second reason is a direct consequence of our axiomatic modeling of optimality: if we are in a regime as in Section~\ref{sec:heuristic-int-dep}, then other code generation strategies for the integration loop might result in larger performance improvements. This is the primary reason we are studying \texttt{Hyperelasticity}, a form characterized by extensive tensor algebra. Nevertheless, as we will see, we found \texttt{cfO2} to be the best scheme also in this circumstance.

A second observation concerns the triggering of pre-evaluation in \texttt{cfO2}. The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} means that pre-evaluation was found unprofitable. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends to not be performed. This is a consequence of the larger increase factor, $\iota$, which we defined in Section~\ref{sec:op_count}.

Also note some other minor things. The lack of improvements in all code variants with $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. The slow downs experienced with \texttt{quad} should be imputable to the presence of indirection arrays in the generated code; when the tables are dense, these eventually break vectorization for a marginal gain due to avoiding iteration over isolated zero-valued columns. A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of \texttt{Hyperelasticity}). Finally, note that the bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.

\paragraph{Mass}
\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix .}\label{fig:mass}
\end{figure}
Apart from a very small number of points in our exploration space, from analysis of Figure~\ref{fig:mass} we can infer how \texttt{cfO2} represents the best code variant among the tested. This suggests that our definition of optimality is consistent in what is the simplest possible bilinear form we can test. As we increase the complexity, by adding more coefficients or by varying the polynomial orders, \texttt{cfO2} is always capable of generating optimal loop nests. The biggest exception is $[q=4, p=4, nf=1]$ in which the save on operations over zero-valued entries in tensor mode is considerably higher than the save in flops in \texttt{cfO1} and \texttt{cfO2}.

\paragraph{Helmholtz}
\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation.}\label{fig:helmholtz}
\end{figure}
Figure~\ref{fig:helmholtz} shows the performance improvements achieved by various code variants over the original code can be significant. It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in most of the cases. One can also appreciate the evidence that the relative order of the employed function spaces plays a key role in the code synthesis choice. Take, for example, the case $[q=3, p=1, nf=2]$. The adoption of pre-evaluation makes a tremendous speed up be achievable by $cfO2$ over all other variants. On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceed the threshold within which tensor mode is applied), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.

@@

\paragraph{Hyperelasticity}
\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model}\label{fig:hyperelasticity}
\end{figure}
In the hyperelastic model experiments, pre-evaluation is never triggered by \texttt{cfO2}. This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction. Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques to this purpose. Results are displayed in Figure~\ref{fig:hyperelasticity}. \texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions. \texttt{ufls} performs generally well and exhibits significant speed ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.

\section{Conclusions}
\label{sec:conclusions}
With this research we have made a first step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration. The results are extremely encouraging, suggesting our model applies to a variety of contexts. We have discussed the conditions under which our optimality definition becomes weaker. An open problem is understanding how to refine this definition to include outer loops. This will probably require studying and exploiting properties of different differential operators. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the high order function spaces.

>>>>

\subsection{Performance of Forms}
\label{sec:perf-results-forms}
% 52 test cases per form, so 52*4 = 208 test cases in total. In 7, there is a non-marginal slow down in cfO2.

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.03$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low-level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} tends to slightly outperform \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries. In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.

A second observation concerns the "triggering" of pre-evaluation in \texttt{cfO2}. The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

Also note some other minor things.
\begin{itemize}
\item The lack of improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over all zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens in the case of the hyperelastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

\paragraph{Mass}
\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix .}\label{fig:mass}
\end{figure}
From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} represents the best variant among the ones tested, apart from the few points previously discussed. This suggests that the optimality model holds in these simple bilinear forms; that is, \texttt{cfO2} generates loop nests that are optimal from both a theoretical and practical viewpoints. It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in many of the test cases, especially as $nf \geq 2$.

\paragraph{Helmholtz}
\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation.}\label{fig:helmholtz}
\end{figure}
Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant. We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice. Take, for example, the case $[q=3, p=1, nf=2]$. The adoption of pre-evaluation makes a tremendous speed up be achievable by $cfO2$ over the competitors. On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.

@@

\paragraph{Hyperelasticity}
\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model}\label{fig:hyperelasticity}
\end{figure}
In the hyperelastic model experiments, pre-evaluation is never triggered by \texttt{cfO2}. This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction. Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques. Results are displayed in Figure~\ref{fig:hyperelasticity}. \texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions. \texttt{ufls} performs generally well and exhibits significant speed ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.

\section{Conclusions}
\label{sec:conclusions}
With this research we have made a first step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration. The results are extremely encouraging, suggesting our model applies to a variety of contexts. We have discussed the conditions under which the optimality model becomes weaker. An open problem is understanding how to refine this model to include outer loops. This will probably require exploiting properties of differential operators. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces.
